Modelling BCG vaccination in the UK

### 2019-12-16 rmds/05-evidence-policy-change.Rmd
rough conversion of sutherland updates

The Bacillus Calmette–Guérin (BCG) vaccine remains the only licensed vaccine for use against Tuberculosis (TB), although its use globally is controversial due to evidence of variable effectiveness,[@Mangtani2014a] and waning protection 10-15 years after vaccination.[@Abubakar2013] Global usage of the BCG varies between no vaccination, universal vaccination, and high-risk group vaccination and may target either neonates or school-aged children.[@Pilger2012b; @Zwerling2011a] The World Health Organization (WHO) recommends vaccination for all neonates as early as possible after birth in high burden settings, with vaccination in low burden settings being dependent on the country specific epidemiology of TB.[@WHO2017] This recommendation is based on the strong evidence that the BCG is highly protective in children,[@Rodrigues1993; @Colditz1994] whilst its effectiveness has been shown to vary with latitude when given later in life (see Chapter \@ref(background)).[@Mangtani2014]

In England and Wales, universal school-aged (at 13-14 years old) vaccination was introduced after a MRC trial in the 1950s estimated BCG's effectiveness at 78% in the ethnic White UK born population.[@Hart1972] The policy remained in place until 2005, when England and Wales changed to targeted vaccination of high risk neonates. The 2005 change in BCG vaccination policy was motivated by evidence of decreased transmission of TB, an increasing proportion of TB cases occurring in the non-UK born,[@PHE2017] and modelling evidence that suggested stopping the BCG schools scheme would have minimal long term effects on incidence rates.[@Sutherland1989] Due to the complex nature of both TB and the BCG vaccine, the ongoing impact of this change in policy is hard to directly estimate, with decision makers relying on expert opinion, evidence from surveillance data, and insight from modelling studies.

In 1987, an assessment of the school-age vaccination program was carried out in England and Wales.[@Sutherland1989] This study was used, combined with a sensitivity analysis of notification rates, as supporting evidence by the JCVI BCG subgroup for the change in vaccination policy.[@JVCIBCG2002; @JVCIBCG2003] This chapter aims to re-evaluate this modelling, and re-estimate the predicted impact of stopping the schools scheme.

@@

I implemented Sutherland et al.’s model for estimating the impact of ending the BCG schools scheme.[@Sutherland1989] This model was based on data from TB notification surveys conducted in 1973, 1978, and 1983.[@Sutherland1987a] These were used to estimate incidence rates, stratified by BCG vaccination status (vaccinated, unvaccinated, and ineligible), in the ethnic White UK born population of England and Wales aged 15-19 years old, 20-24 years old and 25-29 years old. Future incidence rates were forecast by assuming an annual decrease in incidence rates, which was based on historic trends.[@Sutherland1989; @Springett1988] Primary impacts from ending the schools scheme were estimated by calculating the difference in incidence rates between the vaccinated and unvaccinated populations. Additional notifications from TB transmission were then calculated using a transmission chain model. Based on data availability the model used a 5 year timestep. More detail is given in the following sections.

#### Estimating notification rates

The effectiveness of the BCG vaccine was originally estimated by an MRC trial in 1953 at 78% in the United Kingdom.[@Hart1972] As a follow up to this trial members of the MRC bio-statistics unit conducted a series of notification surveys attempting to ascertain any change in effectiveness, as well as acting as an estimate of notification rates across different demographics.[@Sutherland1987a] Notification surveys were carried out at 5 year intervals in 1973, 1978 and 1983; looking at those aged 15-24 years old in England and Wales. For the 1983 survey records of BCG status, Tuberculin status and ethnicity were extracted from the records of notifying physicians and the records of the local health and education authorities. Totals, across the study period, were then aggregated for the following groups: Tuberculin negative and BCG positive, Tuberculin negative and BCG negative, Tuberculin positive and not vaccinated and those who did not participate. These totals were then combined with the population estimates for each cohort at 13 years of age to estimate the ethnic make up of the population, and to construct notification rates for each category. Data were drawn from a range of sources including: Office of National Statistics data; annual local authority returns for total tuberculin test results; BCG vaccinations in the schools scheme; and the Labour force survey (1983).

For 1983 this corresponded with 874 notifications in the ethnic White UK born in England and Wales, between 15-24, with participation at 80%. As the number of Tuberculin negative subjects not given BCG was unreported this was estimated at 1.9% of those vaccinated with the BCG. Allowances were also made for those already vaccinated prior to the schools scheme. See [@Sutherland1987a] for full details of the survey construction and the additional assumptions used to give similar estimates for both the 1973 and 1978 surveys. The findings of these surveys were as follows: in the ethnic White population notification rates had fallen by an annual average of 9% and BCG efficacy had remained high.[@Sutherland1989; @Springett1988]

Evidence suggests that the BCG vaccine has a high efficacy for at least the first 15 years after vaccination, therefore the authors extrapolated from the data on the 20-24 cohort to a theoretical 25-29 year old cohort. Data on the notifications for the first 6 months of the 1983 survey was available and this was then scaled using the ratio of notifications from this age group against the total number of notifications recorded in that year. Population estimates from the 20-24 cohort were adjusted for all causes of mortality (0.34%). Migration was ignored. The tuberculin positive cohort had a sharp decline in the previous two cohorts, therefore it was assumed that this continued. Lastly, the efficacy was estimated as being that seen in the 20-24 cohort but with the same decline in protection seen between the last two cohorts. These assumptions allowed notification rates to be estimated for the 25-29 year old population, resulting in a complete cohort over the projected 15 years of BCG effectiveness.

#### Construction of forward estimates

Based on these estimated notification rates Sutherland et al. then sought to quantify the ongoing risk of developing notified TB, projected forward in time, for both the vaccinated and unvaccinated populations. To construct these estimates several key assumptions, based on the results seen in the previous surveys, were made. Firstly, it was assumed that efficacy was not degrading within the ethnic White population and therefore historic estimates would continue to apply into the future. Additionally, it was assumed that the decay of 9% in notification rates, across all ethnic White populations, would continue indefinitely.

These assumptions allow the notification rates in both the BCG vaccinated and unvaccinated groups to be projected forward in time. By assuming that the schools scheme is responsible for the observed variation between vaccinated and unvaccinated rates the rate of prevented cases can then be estimated. By scaling this against a cohort of 100,000 13 year olds the number of prevented cases over a 15 year period can be projected for each cohort. By dividing the total number in a given cohort by the number of prevented cases the estimated number of vaccines required to prevent a single case in the 15 year period can then be calculated.

To estimate the total number of prevented notifications, for each cohort, in England and Wales the total number receiving the BCG and the coverage of the schools scheme was required. The coverage of the BCG schools scheme was estimated from annual reports of the DHSS and was assumed to be 75% for all years. The number of BCG vaccines given each year was estimated from the DHSS returns for the years 1967 to 1981, it was then taken as 75% of the estimated ethnic White population aged 13 years from 1982-1996, for each 5 year period thereafter it was assumed to be 2.1 million.

Using this information and the projected differences between vaccinated and unvaccinated notification rates allows the number of prevented notifications for each age group to be found for each year. This can then be combined to give the total number of prevented notifications for those aged between 15-29. To understand these estimates, estimates of the projected yearly notifications if the scheme continues are required. These totals were derived from the vaccinated and unvaccinated rates supplemented with similar projections from the tuberculin positive or otherwise ineligible sourced from the 1983 BCG survey.[@Sutherland1987a]

#### Transmission chain model

The TB transmission model is defined as follows (reproduced from Sutherland et al.):

1. Estimate the total expected number of secondary notifications ($T$) arising from any single primary notification using the following,

$$ T = (1-d)^z < 1 $$
Where $d$ is the percentage decay in notification rates, and $z$ is the average interval between the notification of any individual and the notification of the patient who infected them.

2. Relate $T$ to the number of notifications in each generation using the initial generation size ($x$) with the following power series,

$$ T=x+x^2+x^3+ . . = \frac{x}{1-x} $$
3. Estimate the expected average interval between each primary notification and all secondary notifications ($Z$). This is defined to be the sum of time to all notifications, weighted by the fraction in each generation, divided by the sum of all notifications. Mathematically this is,

$$ Z = \frac{xz + x^{2}2z + x^{3}3z + . .}{x+x^2+x^3+ . . }= \frac{z}{1-x} $$

#### Transmission chain model additions

Implementing the model required several assumptions not detailed in [@Sutherland1989]. Firstly, as incidence rates for those ineligible for the BCG schools scheme are not published, I assumed that they were equal to those in the unvaccinated population. In addition, in order to reproduce the distribution of cases over time reported in Sutherland et al. I introduced an additional model step and parameter; the percentage of secondary cases due to a primary case in the first year after activation ($f$). The distribution of secondary cases ($N$) was then modelled by first estimating the number of secondary cases that occurred in the current year and then estimating how many secondary cases occurred 5 years later (due to the 5 year timestep). This iterative process can be summarised as follows,

$$ N^{Current} = P T  f  x $$
$$ N^{Projected} = (PT - N^{Current})(1 - d)^{L-Z} $$ 
$$ N_i^{Total} = N^{Current}_i + N^{Projected}_{i-1} $$

Where $P$ is the number of primary notifications, $i$ is the year of interest, $N^{Current}$ is the number of secondary cases in the current  year, $N^{Projected}$ is the number of secondary cases in the subsequent timestep, and $L$ is the model timestep (set to be 5 years).

I fitted this parameter using least squares to the original estimates of the total notifications due to ending the scheme under several scenarios, for several years. I validated the fitted model by comparing the results with those from the original implementation. Using the mean absolute percentage error, normalised by the orginal estimate.

### Updating model parameter estimates

Incidence rates were included as point estimates in [@Sutherland1989]; in the updated model I included uncertainty in these rates. I estimated notifications for 1973, 1978, and 1983, using published incidence rates and population estimates. Samples were then generated using a Poisson distribution.[@Sutherland1989; @Sutherland1987a] These samples were then used to estimate a distribution of incidence rates. Sutherland et al. assumed a serial interval of 2 years between linked infections. I updated this assumption using, the currently best available estimate, of 1.44 (95% CI 1.29 to 1.63) years.[@Borgdorff2011]

I considered the original assumption of a 9% annual decrease in incidence rates as well as three scenarios based on those considered by the JCVI BCG subgroup:[@JVCIBCG2002; @JVCIBCG2003] these were a 3.9% decrease, a 1.9% decrease, and no change annual in incidence rates. I also estimated the annual decrease in incidence rates in the ethnic White UK born using two proxy measures. The first proxy measure was the annual change in notifications in England and Wales, which was estimated using data from Public Health England (PHE). The standard deviation (SD) of this measure was then calculated using the `prop.test` function in R.[@R] The second proxy used was the annual decrease in the UK born age-specific incidence rates in the English population. These were calculated using notification data from the Enhanced TB surveillance system (ETS) and the June Labour Force Survey (see Chapter \@ref(data)).[@PHE2017] Incidence rates (with SDs) were estimated using the epiR package.[@EpiR] Uncertainty was incorporated by sampling from a normal distribution for both proxy measures. Data collection for the ETS began in 2000, I therefore estimated incidence rates between 1984 and 1999, and for the years between notifications surveys (1974-1977 and 1979-1892), using LEOSS regression for each sample fitted to the incidence rates published in [@Sutherland1989], and the estimated incidence rates from 2000 on-wards. LOESS (locally estimated scatterplot smoothing) is a local regression method.[@FoxSnow:2019] It combines multiple regression models in a k-nearest neighbours meta-model. This approach allows nonlinear trends to be fitted to using a series of linear models. For years prior to 1973 the annual decreases were assumed to be the mean of the previous 3 years of data. For both proxy measures the annual decreases in incidence rates post 2016 were assumed to be the average of the estimates in 2013, 2014, and 2015.

@@

For each scenario, I ran the model until 2028 with 10,000 parameter samples. I tested the difference between scenarios using the Mann-Whitney test for both the number of vaccines needed to prevent a single case in 15 years after vaccination for a cohort aged 13-14 years old at vaccination and the total number of additional notifications caused by ending the BCG schools scheme. As in [@Sutherland1989] a 15 year time horizon was used with 5 year intervals. The year closest to the year of the change in vaccination policy (2005), which had model estimates, was used as the baseline. The code for this analysis is available online^[Code: https://github.com/seabbs/AssessBCGPolicyChange].

@@

### Model validation

@@

```{r plot-annual-change, fig.scap = "Annual percentage change in ethnic White UK born incidence rates for those aged 15-19, 20-24, and 25-29 years old under different scenarios.", fig.cap = "Annual percentage change in ethnic White UK born incidence rates for those aged 15-19, 20-24, and 25-29 years old under different scenarios. For the notification and incidence rate scenarios each line represents the median of 10,000 parameter samples.", out.extra = ""}
knitr::include_graphics("chapters/evidence-policy-change/plot-annual-change-1.png")
```

@@

I found that updating parameter values, and incorporating uncertainty, did not alter the number of vaccines required to prevent a single notification within 15 years in a cohort vaccinated at school-age, when the annual decrease in TB incidence rates was assumed to be 9% (Figure \@ref(fig:graph-vacs-to-prev); Table \@ref(tab:vac-prevent-tab)). However, the updated estimate had a wide range (`r vac_2004_9` vaccines required in 2004). As the assumed annual decrease in incidence rates was reduced the number of vaccines required to prevent a single notification also reduced; an estimated `r vac_2004_1` vaccines were required to prevent a case in 2004 when the annual decrease was assumed to be 1.9%. Estimates of the number of vaccines required to prevent a notification were comparable but not equivalent when the annual decrease was estimated using notifications (`r vac_2004_nots`, P: `r p_not_vs_1_2004`) and age-specific incidence rates (`r vac_2004_inc`, P: `r p_inc_vs_1_2004`). The estimate using incidence rates had a high degree of uncertainty (Figure \@ref(fig:graph-vacs-to-prev); Table \@ref(tab:vac-prevent-tab)). The number of vaccines required increased slightly over time with `r vac_prevent_1_2009` required in 2009, `r vac_prevent_1_2014` required in 2014, and `r vac_prevent_1_2019` required in 2019 when an annual decrease of 1.9% in incidence rates was assumed.

```{r graph-vacs-to-prev, fig.scap = "Vaccines required in a cohort of those vaccinated at school-age to prevent a single case of TB within 15 years of vaccination in 2004, 2009, 2014, or 2019.", fig.cap = "Vaccines required in a cohort of those vaccinated at school-age to prevent a single case of TB within 15 years of vaccination in 2004, 2009, 2014, or 2019. The years presented were dictated by the 5 year timestep of the model. The percentage annual decrease scenarios considered were based on those considered by the JCVI BCG subgroup, with the addition of a scenario using aggregate notification data and a scenario using estimates of age-specific incidence rates in the UK born. Each boxplot summarises the output of 10,000 model simulations for each scenario. Outliers have been omitted for clarity.", out.extra = ""}
knitr::include_graphics("chapters/evidence-policy-change/graph-vacs-to-prev-1.png")
```

@@

I found that updating parameter values, and incorporating uncertainty, did not alter the average annual additional notifications from stopping the BCG schools scheme when the annual decrease was assumed to be 9% (Figure \@ref(fig:graph-avg-add-not); Table \@ref(tab:annual-prevent-tab)). There was a large degree of uncertainty in this estimate with `r avg_annual_2001_9` notifications prevented annually if vaccination was stopped in 2001. As the assumed annual decrease in incidence rates was reduced the annual number of additional notifications prevented increased with `r avg_annual_2001_1` notifications prevented annually when the annual decrease was assumed to be 1.9% and vaccination stopping in 2001. There was some evidence that the average annual number of notifications prevented was greater when the annual decrease was estimated using notifications (`r avg_annual_2001_nots`, P: `r a_not_vs_1_2001`) and age-specific incidence rates (`r avg_annual_2001_inc`, P: `r a_inc_vs_1_2001`), compared to an assumed annual decrease of 1.9% (Figure \@ref(fig:graph-avg-add-not); Table \@ref(tab:annual-prevent-tab)). The estimate made using incidence rates again had a high degree of uncertainty. When an annual decrease of 1.9% was assumed the number of notifications prevented annually reduced with time: `r avg_annual_1_2006` from ending vaccination in 2006; `r avg_annual_1_2011` from ending vaccination in 2011, and `r avg_annual_1_2016` from ending vaccination in 2016.

```{r graph-avg-add-not, fig.scap = "Annual additional notifications in 15-29 year olds from stopping the BCG schools scheme in 2001, 2006, 2011, and 2016 until 2028.", fig.cap = "Annual additional notifications in 15-29 year olds from stopping the BCG schools scheme in 2001, 2006, 2011, and 2016 until 2028. The years presented were dictated by the 5 year timestep of the model. The percentage annual decrease scenarios considered were based on those considered by the JCVI BCG subgroup, with the addition of a scenario using aggregate notification data and a scenario using estimates of age-specific incidence rates in the UK born. Each boxplot summarises the output of 10,000 model simulations for each scenario. Outliers have been omitted for clarity.", out.extra = ""}
knitr::include_graphics("chapters/evidence-policy-change/graph-avg-add-nots-1.png")
```

```{r annual-prevent-tab}
  read_formated("tab-annual_nots_prevented") %>% 
  kable(
    caption = "The median number (with the 2.5\\% and 97.5\\% quantiles) of additional annual notifications due to ending the BCG schools scheme in selected years.", booktabs = TRUE) %>% 
  landscape() %>% 
  kable_styling(latex_options = "hold_position")  %>% 
  kable_styling(font_size = 8) %>% 
  column_spec(2, width = "2cm")
```

@@

* I found a 1.9% annual decrease in TB incidence rates best matched estimates from notifications.

* I estimate that 1600 (2.5-97.5% quantiles (Q): 1300 - 2100) vaccines would have been required to prevent a single notification in 2004.

>>>>

The Bacillus Calmette–Guérin (BCG) vaccine remains the only licensed vaccine for use against Tuberculosis (TB), although its use globally is controversial due to evidence of variable effectiveness,[@Mangtani2014a] and waning protection 10-15 years after vaccination.[@Abubakar2013] Global usage of the BCG varies between no vaccination, universal vaccination, and high-risk group vaccination and may target either neonates or school-aged children.[@Pilger2012b; @Zwerling2011a] The World Health Organization (WHO) recommends vaccination for all neonates as early as possible after birth in high burden settings, with vaccination in low burden settings being dependent on the country specific epidemiology of TB.[@WHO2017] This recommendation is based on the strong evidence that the BCG is highly protective in children,[@Rodrigues1993; @Colditz1994] whilst its effectiveness has been shown to vary with latitude when given later in life.[@Mangtani2014]

In England and Wales, universal school-aged (at 13-14 years old) vaccination (hereafter referred to as the BCG schools scheme) was introduced after a MRC trial in the 1950s estimated BCG's effectiveness at 80% in the ethnic White UK born population.[@Hart1972] The policy remained in place until 2005, when England and Wales changed to targeted vaccination of high-risk neonates. The 2005 change in BCG vaccination policy was motivated by evidence of decreased transmission of TB, an increasing proportion of TB cases occurring in the non-UK born,[@PHE2017] and modelling evidence that suggested stopping the BCG schools scheme would have minimal long term effects on incidence rates.[@Sutherland1989] Due to the complex nature of both TB and the BCG vaccine, the ongoing impact of this change in policy is hard to directly estimate, with decision makers relying on expert opinion, evidence from surveillance data, and insight from modelling studies.

In 1987, an assessment of the school-age vaccination program was carried out in England and Wales.[@Sutherland1989] This study was used, combined with a sensitivity analysis of notification rates, as supporting evidence by the Joint Committee on Vaccination and Immunisation (JCVI) BCG subgroup for the change in vaccination policy.[@JVCIBCG2002; @JVCIBCG2003] This chapter aims to re-evaluate this modelling, and re-estimate the predicted impact of stopping the schools scheme. Re-evaluating this work allows for the strength of the evidence used in decision making to be assessed and may highlight any issues with the approach used.

@@

I implemented, and updated, Sutherland et al.’s model for estimating the impact of ending the BCG schools scheme, which is outlined briefly below.[@Sutherland1989] This model was based on data from TB notification surveys conducted in 1973, 1978, and 1983.[@Sutherland1987a] These were used to estimate incidence rates, stratified by BCG vaccination status, in the ethnic White UK born population of England and Wales aged 15-19 years old, 20-24 years old and 25-29 years old. Future incidence rates were forecast by assuming an annual decrease in incidence rates, which was based on historic trends.[@Sutherland1989; @Springett1988] Primary impacts from ending the schools scheme, including the number of vaccines required to prevent a single notification, were estimated by calculating the difference in incidence rates between the vaccinated and unvaccinated populations. Additional notifications from TB transmission were then calculated using a transmission chain model and combined with the primary impact estimates, to calculate the number of annual additional notifications due to ending BCG vaccination. Based on data availability, the model used a 5-year time step.

#### Original estimation of notification rates

The effectiveness of the BCG vaccine was originally estimated by an MRC trial in 1953 at 78% in the United Kingdom.[@Hart1972] As a follow up to this trial members of the MRC bio-statistics unit conducted a series of notification surveys attempting to ascertain any change in effectiveness, as well as acting as an estimate of notification rates across different demographics.[@Sutherland1987a] Surveys of those aged 15-24 years were carried out at 5-year intervals in 1973, 1978 and 1983 in England and Wales. For the 1983 survey records of BCG status, Tuberculin status and ethnicity were extracted from the records of notifying physicians and the records of the local health and education authorities. Total notifications across the study period were then aggregated for the following groups: Tuberculin negative and BCG vaccinated, Tuberculin negative and BCG unvaccinated, Tuberculin positive and not vaccinated and those who did not participate. These totals were then combined with the population estimates for each cohort at 13 years of age to estimate the ethnic make up of the population, and to construct notification rates for each category. Data were drawn from a range of sources including: Office of National Statistics data; annual local authority returns for total Tuberculin test results; BCG vaccinations in the schools scheme; and the Labour force survey (1983).

For 1983, there were 874 notifications recorded in 15-24 year old ethnic White UK born persons in England and Wales; survey participation was 80%. As the number of Tuberculin negative subjects not given BCG was unreported the number of notifications was estimated at 1.9% of those vaccinated with the BCG.[@Sutherland1987a] See [@Sutherland1987a] for full details of the survey and the additional assumptions used to give similar estimates for both the 1973 and 1978 surveys. The findings of these surveys were as follows: in the ethnic White population notification rates had fallen by an annual average of 9% and BCG efficacy had remained high.[@Sutherland1989; @Springett1988]

Evidence suggests that the BCG vaccine has a high efficacy for at least the first 15 years after vaccination, therefore Sutherland et al. extrapolated from the data on the 20-24 cohort to a theoretical 25-29 year old cohort. Data on the notifications in 25-29 year olds was available for the first 6 months of the 1983 survey and this was then scaled up to a yearly estimate using the ratio of notifications from this age group against the total number of notifications recorded in that year. Population estimates for the 25-29 year old cohort were based on data from the 20-24 cohort adjusted for all causes of mortality (0.34%). Migration was ignored. The Tuberculin positive cohort had a sharp decline in the previous two age cohorts, therefore it was assumed that this continued. Lastly, the efficacy was estimated as being that seen in the 20-24 cohort but with the same decline in protection seen between the last two cohorts. These assumptions allowed notification rates to be estimated for the 25-29 year old population, resulting in a complete cohort over the projected 15 years of BCG effectiveness.

#### Original construction of forward estimates

Based on these estimated notification rates, Sutherland et al. then sought to quantify the ongoing risk of developing notified TB, projected forward in time, for both the vaccinated and unvaccinated populations. To construct these estimates several key assumptions, based on the results seen in the previous surveys, were made. Firstly, it was assumed that efficacy was not degrading within the ethnic White population and therefore historic estimates would continue to apply into the future. Additionally, it was assumed that the annual decay of 9% in notification rates, across all ethnic White populations, would continue indefinitely.

These assumptions allowed the notification rates in both the BCG vaccinated and unvaccinated groups to be projected forward in time. By assuming that the schools scheme is responsible for the observed variation between vaccinated and unvaccinated rates the rate of prevented cases can then be estimated. By scaling this against a cohort of 100,000 13 year old individuals, the number of prevented cases over a 15-year period can be projected for each cohort. By dividing the total number in a given cohort by the number of prevented cases the estimated number of vaccines required to prevent a single case in the 15-year period can then be calculated.

To estimate the total number of prevented notifications, for each cohort, in England and Wales the total number receiving the BCG and the coverage of the schools scheme was required. The coverage of the BCG schools scheme was estimated from annual reports of the Department of Health and Social Security (DHSS) and was assumed to be 75% for all years. The number of BCG vaccines given each year was estimated from the DHSS returns for the years 1967 to 1981, it was then taken as 75% of the estimated ethnic White population aged 13 years from 1982-1996, for each 5-year period thereafter it was assumed to be 2.1 million.

Using the data on BCG coverage, the number of vaccines given each year, and the projected differences between vaccinated and unvaccinated notification rates allowed the number of prevented notifications, due to vaccination, for each age group to be found for each year. These estimates can then be used to give the total number of prevented notifications for those aged between 15-29 years. To understand these estimates, estimates of the projected yearly notifications if the scheme continues were required. These totals were derived from the vaccinated and unvaccinated rates supplemented with similar projections from the Tuberculin positive or otherwise ineligible sourced from the 1983 BCG survey.[@Sutherland1987a]

#### Original transmission chain model

Sutherland et al. defined their TB transmission model as follows:

1. The total expected number of secondary notifications ($T$) arising from any single primary notification was estimated as,

\[ T = (1-d)^z < 1 \]

Where $d$ is the percentage annual decay in notification rates, and $z$ is the average interval between the notification of any individual and the notification of the patient who infected them.

2. The total expected number of secondary notifications arising from any single primary notification ($T$) is related to the number of notifications in each generation using the relative generation size ($x$) with the following power series,

\[ T=x+x^2+x^3+ . . = \frac{x}{1-x} \]

3. The expected average interval between each primary notification and all secondary notifications ($Z$) is defined to be the sum of time to all notifications, weighted by the fraction in each generation, divided by the sum of all notifications.

\[ Z = \frac{xz + x^{2}2z + x^{3}3z + . .}{x+x^2+x^3+ . . }= \frac{z}{1-x} \]

Both 2. and 3. are only valid when $x < 1$.

#### Updating the transmission chain model

If we assume a constant decay rate of $d$% per year and that the next generation of secondary cases are notified $z$ years after the person who infected them, then 1 notification in year 0 will result in $(1-d)^z$ secondary notifications $z$ years later. This is therefore the relative generation size ($x$), not the *total* number of secondary notifications ($T$), as (incorrectly) stated by Sutherland et al. Therefore, to correct this I revised 1. to the following in our updated model,

\[ x = (1-d)^z < 1 \]

The Sutherland et al. model required several additional assumptions.[@Sutherland1989] Firstly, as incidence rates for those ineligible for the BCG schools scheme are not published, I assumed that they were equal to those in the unvaccinated population. In addition, in order to reproduce the distribution of cases annually (rather than by generation or overall) reported in Sutherland et al. I introduced an additional model step and parameter; the proportion of secondary cases in the first generation that occurred in the first year ($f$). This modelling step is only required to reproduce the final table from [@Sutherland1989] and does not impact estimates of the impact of ending the BCG schools scheme. It is included only for validation purposes.

The annual distribution of secondary notifications ($N$) was modelled by first estimating the number of secondary notifications that occurred in the current year ($i$) due to primary notifications in that year ($N^{Current}$) and then estimating how many secondary notifications occurred 5 years later ($N^{Projected}$). $N^{Current}$ was estimated using the number of primary notifications ($P$) multiplied by the number of total expected number of secondary notifications per primary notification ($T$), the proportion of secondary cases in the first generation that occurred in the first year ($f$), and the relative size of the first generation ($x$). $N^{Projected}$ was then estimated by assuming that it was equal to the number of secondary notifications, minus notifications occurring in the first year, that occurred $Z$ (the expected average interval between each primary notification and all secondary notifications) years ago. As this used the overall number of notifications from the previous time step a decay of $(1 - d)^{5-Z}$ was applied. This approach can be summarised as follows,

\[ N^{Current} = P T  f  x \]
\[ N^{Projected} = (PT - N^{Current})(1 - d)^{5-Z} \] 
\[ N_i = N^{Current}_i + N^{Projected}_{i-1} \]

I fitted the proportion of secondary cases in the first generation that occurred in the first year ($f$) using least squares to the original estimates of the total notifications due to ending the scheme under several scenarios, for several years. I validated the fitted model by comparing the results with those from the original implementation using the mean absolute percentage error, normalised by the original estimate, as the performance metric.

### Updating model parameter estimates and incorporating parameter uncertainty

Incidence rates were included as point estimates in [@Sutherland1989]; in the updated model I included uncertainty in these rates. I did this by first estimating notifications for 1973, 1978, and 1983, using published incidence rates and population estimates. Samples were then generated using a Poisson distribution.[@Sutherland1989; @Sutherland1987a] These samples were then used to estimate a distribution of incidence rates to replace the point estimates used in the original analysis. Sutherland et al. assumed a serial interval ($z$) of 2 years between linked notifications. Using a newly available literature source I updated this assumption with an estimate of 1.44 (95% CI 1.29 to 1.63) years.[@Borgdorff2011]

I considered the original assumption of a 9% annual decrease in incidence rates as well as three scenarios based on those considered by the JCVI BCG subgroup:[@JVCIBCG2002; @JVCIBCG2003] these were a 3.9% decrease, a 1.9% decrease, and no change annual in incidence rates. Data on the annual decrease in incidence rates in the ethnic White UK population were not available so I used two proxy measures. The first proxy measure was the annual change in notifications in England and Wales, which was estimated using data from Public Health England (PHE). The standard deviation of this measure was then calculated using the `prop.test` function in R.[@R] The second proxy used was the annual decrease in the UK born age-specific incidence rates in the English population. These were calculated using notification data from the Enhanced TB surveillance system (ETS) and the June Labour Force Survey.[@PHE2017] Incidence rates were estimated using the epiR package.[@EpiR] Uncertainty was incorporated by sampling from a normal distribution for both proxy measures. Data collection for the ETS began in 2000 and prior to this notification data was only available in years with notifications surveys (1973, 1978, and 1983). I therefore estimated incidence rates between 1984 and 1999, and for the years between notifications surveys (1974-1977 and 1979-1982), using locally estimated scatter plot smoothing (LOESS) regression fitted to incidence rates published in [11] and the estimated incidence rates from 2000 on-wards. LOESS is a local regression method that combines multiple regression models in a k-nearest neighbours meta-model.[@FoxSnow:2019] This approach allows nonlinear trends to be fitted using a series of linear models. For years prior to 1973 the annual decreases were assumed to be the mean of the annual decreases from the previous 3 years. For both proxy measures the annual decreases in incidence rates post 2016 were assumed to be the average of the estimates in 2013-2015.

@@

For each scenario, I ran the model for 69 years (1969-2028) with 10,000 parameter samples. I tested the difference between scenarios using the Mann-Whitney test for the number of vaccines needed to prevent a single case in 15 years after vaccination for a cohort aged 13-14 years old at vaccination. As in [@Sutherland1989] a 15-year time horizon was used with 5-year intervals. The year closest to the year of the change in vaccination policy (2005), which had model estimates, was used as the baseline. The code for this analysis is available online^[Code: https://github.com/seabbs/AssessBCGPolicyChange].

@@

@@

```{r plot-annual-change, fig.scap = "Annual percentage change in ethnic White UK born incidence rates for those aged 15-19, 20-24, and 25-29 years old under different scenarios.", fig.cap = "Annual percentage change in ethnic White UK born incidence rates for those aged 15-19, 20-24, and 25-29 years old under different scenarios. For the notification and incidence rate scenarios each line represents the median of 10,000 parameter samples.", out.extra = ""}
knitr::include_graphics("chapters/evidence-policy-change/fig-1.tiff")
```

@@

I found that incorporating uncertainty, did not alter the number of vaccines required to prevent a single notification within 15 years in a cohort vaccinated at school-age, when the annual decrease in incidence rates was assumed to be 9% (Figure \@ref(fig:graph-vacs-to-prev); Table \@ref(tab:vac-prevent-tab)). However, the updated estimate had a wide range (`r vac_2004_9` vaccines required in 2004). As the assumed annual decrease in incidence rates was reduced the number of vaccines required to prevent a single notification also reduced. Assuming an annual decrease of 1.9% (one of the scenarios evaluated by the JCVI) resulted in an estimate of `r vac_2004_1` vaccines required to a prevent a single notification in 2004. This assumption was the most comparable, although not equivalent, to estimates derived using notifications (`r vac_2004_nots`, P: `r p_not_vs_1_2004`) and age-specific incidence rates (`r vac_2004_inc`, P: `r p_inc_vs_1_2004`). The estimate using incidence rates had a high degree of uncertainty (Figure \@ref(fig:graph-vacs-to-prev); Table \@ref(tab:vac-prevent-tab)). The number of vaccines required increased slightly over time with `r vac_prevent_1_2009` required in 2009, `r vac_prevent_1_2014` required in 2014, and `r vac_prevent_1_2019` required in 2019 when an annual decrease of 1.9% in incidence rates was assumed.

```{r graph-vacs-to-prev, fig.scap = "Vaccines required in a cohort of those vaccinated at school-age to prevent a single case of Tuberculosis within 15 years of vaccination in 2004, 2009, 2014, or 2019.", fig.cap = "Vaccines required in a cohort of those vaccinated at school-age to prevent a single case of Tuberculosis within 15 years of vaccination in 2004, 2009, 2014, or 2019. The years presented were dictated by the 5-year timestep of the model. The percentage annual decrease scenarios considered were based on those considered by the JCVI BCG subgroup, with the addition of a scenario using aggregate notification data and a scenario using estimates of age-specific incidence rates in the UK born. Each boxplot summarises the output of 10,000 model simulations for each scenario.", out.extra = ""}
knitr::include_graphics("chapters/evidence-policy-change/fig-2.png")
```

@@

I found that updating parameter values, and incorporating uncertainty, did not alter the average annual primary additional notifications from stopping the BCG schools scheme when the annual decrease was assumed to be 9%. However, when these changes were combined with the updated transmission model we found that the impact of ending BCG vaccination was greater than previously reported with an increase in the number of estimated cases due to onwards transmission (Figure \@ref(fig:graph-avg-add-not); Table \@ref(tab:annual-prevent-tab)). These estimates were uncertain with `r avg_annual_2001_9` additional annual notifications if vaccination was stopped in 2001. As the assumed annual decrease in incidence rates was reduced the annual number of additional notifications increased with `r avg_annual_2001_1` notifications when the annual decrease was assumed to be 1.9% and vaccination stopped in 2001. The number of annual notifications reduced with time: `r avg_annual_1_2006` from ending vaccination in 2006; `r avg_annual_1_2011` from ending vaccination in 2011, and `r avg_annual_1_2016` from ending vaccination in 2016 (Figure \@ref(fig:graph-avg-add-not); Table \@ref(tab:annual-prevent-tab)).

```{r graph-avg-add-not, fig.scap = "Annual additional (stratified into primary and secondary) notifications in 15-29 year olds from stopping the BCG schools scheme in 2006, and 2011 until 2028.", fig.cap = "Annual additional (stratified into primary and secondary) notifications in 15-29 year olds from stopping the BCG schools scheme in 2006, and 2011 until 2028. The years presented were dictated by the 5-year timestep of the model. The percentage annual decrease scenarios considered were based on those considered by the JCVI BCG subgroup. Data based scenarios and the JCVI 0% decrease scenario were not presented here as the updated transmission model could not support these scenarios. Each boxplot summarises the output of 10,000 model simulations for each scenario. Secondary notifications are reported assuming they occurred in the same year as the primary notifications that caused them.", out.extra = ""}
knitr::include_graphics("chapters/evidence-policy-change/fig-3.tiff")
```

```{r annual-prevent-tab}
read_formated("tab-annual_nots_prevented") %>% 
  kable(
    caption = "The median number (with the 2.5\\% and 97.5\\% quantiles) of additional annual notifications due to ending the BCG schools scheme in selected years.", booktabs = TRUE) %>% 
  landscape() %>% 
  kable_styling(latex_options = "hold_position")  %>% 
  kable_styling(font_size = 10) %>% 
  column_spec(2, width = "2cm")
```

@@

* This chapter reevaluates a key piece of the quantitative evidence used to motivate the change in BCG vaccination policy in 2005, correcting a methodological flaw in the original model.
